---
title: "A `maps` quirk and a `tmap` solution"
author: "Will Lowe & Asya Magazinnik"
date: "2/8/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For those who are interested, this tutorial walks through the buggy features of the
`maps` package, then provides a better alternative. The important thing to know 
about using the `map()` function in `maps` is to loop over regions to color them 
one by one rather than providing a vector of colors in one function call. We'll 
first illlustrate why. 

## `maps`

Let's load a dataset containing the colors we wish to use for each 
county, identified by FIPS code. 

```{r}
load("data/county_data_full.RData")
head(final)
```

We'll need to merge this dataset with the mapping data used by the `maps` library, by FIPS code.  

```{r}
library(maps)
library(ggmap)

cf <- county.fips # from maps library
names(cf) <- c("FIPS", "name")
toplot <- merge(cf, final, by = "FIPS", all.x = TRUE)
head(toplot)
```

Now let's try making a map by providing a vector of colors to one map call. 

```{r fig.align = 'center'}
map(database = "county", col = toplot$cols1, lty = 0, fill = TRUE)  
```

And let's try an alternative approach: activating an empty map and looping over counties
to color them one by one. 

```{r fig.align = 'center', eval=TRUE}
map(database = "county", lty = 0) # activate empty map
for(i in 1:nrow(toplot)) {
  map(database = "county", regions = toplot$name[i], col = toplot$cols1[i], 
      fill = TRUE, add = TRUE, lty = 0)
}
```

We see the same overall patterns, but if you look closely, there are subtle differences 
between the two maps (look at the Northwest and Northeast tips of the map). Which 
one is right? 

The answer lies in the fact that we used the `county.fips` dataset as a base
for the color vector, but the `map()` function call uses a different dataset
for its geography. We can take a look at it using the `map_data()` function. 

```{r}
geog <- map_data("county")
head(geog)
```

It turns out that the regions in this dataset don't align perfectly with the
`county.fips` dataset. 

```{r}
length(unique(paste(geog$region, geog$subregion))) # county dataset
length(cf$name) # fips dataset
```

Whoops. It turns out these also aren't in the same order, which matters.
Why didn't we get an error if we weren't using a color
vector of the appropriate length? Because, as the documentation will tell you,
it cycles through the vector given to `col`; if it stops short of the
regions in the data, it goes back to the first position in the vector. 

We can still use our first approach, but we have to use the `regions` argument
to explicitly tell it which counties align with our color vector. To do so, 
we'll have to left-join our data onto the geography dataset. 

```{r}
geog <- unique(map_data("county")[,c("group", "region", "subregion")])
geog$name <- paste(geog$region, geog$subregion, sep=",")
merged <- dplyr::left_join(geog, toplot, by="name")
```

Now let's see if we can get the same map as looping. 

```{r eval=FALSE, tidy=FALSE}
map(database = "county", regions = merged$name, col = merged$cols1, 
    lty = 0, fill = TRUE)  
```

Not so fast. It returns the error message: 

```{r eval=FALSE, tidy=TRUE}
Error in grep(regexp, nam, ignore.case = TRUE, perl = (length(patterns) <  : 
  invalid regular expression '(^alabama,autauga)|(^alabama,baldwin)|(^alabama,barbour)|
(^alabama,bibb)|(^alabama,blount)|(^alabama,bullock)|(^alabama,butler)|
(^alabama,calhoun)|(^alabama,chambers)|(^alabama,cherokee)|(^alabama,chilton)|
(^alabama,choctaw)|(^alabama,clarke)|(^alabama,clay)|(^alabama,cleburne)|
(^alabama,coffee)|(^alabama,colbert)|(^alabama,conecuh)|(^alabama,coosa)|
(^alabama,covington)|(^alabama,crenshaw)|(^alabama,cullman)|(^alabama,dale)|
(^alabama,dallas)|(^alabama,dekalb)|(^alabama,elmore)|(^alabama,escambia)|
(^alabama,etowah)|(^alabama,fayette)|(^alabama,franklin)|(^alabama,geneva)|
(^alabama,greene)|(^alabama,hale)|(^alabama,henry)|(^alabama,houston)|(^alabama,jackson)|
(^alabama,jefferson)|(^alabama,lamar)|(^alabama,lauderdale)|(^alabama,lawrence)|
(^alabama,lee)|(^alabama,limestone)|(^alabama,lowndes)|(^alabama,macon)|
(^alabama,madison)|(^alabama,marengo)|(^alabama,marion)|(^alabama,marshall)|
(^alabama,mobile)|(^alabama,monroe)|(^alabama,montgomery)|(^alabama'
```

What now? The issue is that the `map` function's source code uses the `grep` command,
which can only take an argument up to a certain length; all the counties we
want to plot turn out to be too many. If we do it with the first 2500 observations,
we get the same results as looping. 

```{r fig.align = 'center'}
# vector approach
map(database = "county", regions = merged$name[1:2500], col = merged$cols1[1:2500], 
    lty = 0, fill = TRUE) 

# loop approach
map(database = "county", lty = 0) # activate empty map
for(i in 1:2500) {
  map(database = "county", regions = toplot$name[i], col = toplot$cols1[i], 
      fill = TRUE, add = TRUE, lty = 0)
} 
```

The bottom line? You should loop. But that's still slow and clunky, and it's 
difficult to provide your own geographies. A better approach is `tmap`; a 
quick illustration of how it works follows. 

## `tmap`

First, download and unpack some county shapefiles from the the census.  These are 1:20M scale
so the plots go relatively quickly.
```{r}
loc <- file.path('data', "counties1to20m.zip")
download.file("http://www2.census.gov/geo/tiger/GENZ2015/shp/cb_2015_us_county_20m.zip", 
              loc)
unzip(loc, exdir='data')
```

Get the plotting library and its corresponding data manipulation library
```{r}
library(tmap)
library(tmaptools) # for the shape file reader
```

Read the shapefile and turn it into a `SpatialPolygonsDataFrame` which 
behaves mostly like a `data.frame` but has spatial polygons associated with the rows.
```{r}
cts <- read_shape(file.path('data', "cb_2015_us_county_20m.shp"))
```

Create a proper FIPS variable from the two components that the Census provides
```{r}
cts$FIPS <- as.numeric(paste0(cts$STATEFP, cts$COUNTYFP))
```

Now to merge our metadata to the county shapes.  This won't work first time...
```{r}
append_data(cts, final, key.shp = "FIPS", key.data = "FIPS")
```
OK, so let's take a use the output of `under_coverage()` to remove some 
elements from the shape file.  It tells us the ids and FIPS values of the
offending polygons.
```{r}
cts.smaller <- cts[-under_coverage()$id, ]
```
Now let's try that merge again.
```{r}
cts.merged <- append_data(cts.smaller, final, key.shp="FIPS", key.data="FIPS")
```
That's better.  

Notice that `append_data` always *left joins* on the object holding the polygons.
That's why we had to adjust that file a bit.  

You might have been tempted to use 
`merge` for this, but that would have silently thrown away all the shape information 
that's secreted in `cts`, which would have sunk everything (with a confusing 
error about lacking projection information).  So don't do that.

Now to make a nice plot.  `tmap` has a ggplot style syntax using `+`. Here we
just make a simple chloropleth:
```{r}
tm_shape(cts.merged) + tm_fill("cols1")
```